---
title: "Exponential Distribution Simulation"
author: "Ash Chakraborty"
date: "Wednesday, June 17, 2015"
output:
  pdf_document:
    toc: no
  html_document:
    theme: readable
  word_document: default
---

# PART 1: SIMULATION - Sampling the Exponential Distribution

## Synopsis

Part 1 of this project investigates the _Exponential Distribution_ and compares it to the _Central Limit Theorem_. The exponential distribution is given by the probability distribution function, $P(x)=\lambda e^{-\lambda x}$. We also know that it has a mean, $\mu = 1/\lambda$, and variance, $\sigma^{2}=1/\lambda^{2}$. This report will conduct an appropriate number of simulations on this distribution with a sample size of 40 exponentials by generating a sampling distribution of sample means. The consequences of this sampling distribution will be evaluated for adherence to the central limit theorem.  

## Exponential Distribution

We are to work with a sample size, n = 40 with a rate parameter, $\lambda = 0.2$. We first generate a sample of 40 random exponentials, and look at its central tendencies:
```{r global_options, echo=TRUE, warning=FALSE, message=FALSE, tidy=TRUE, highlight=TRUE, fig.path="/figures", fig.width=9}
library(PerformanceAnalytics)
library(ggplot2)
library(gridExtra)
set.seed(123)
sample <- rexp(40, rate=0.2)
```  

# Simulating 1000 Trials

In order to understand the properties of the distribution of the mean of 40 exponentials, we conduct a 1000 simulations, and extract the mean of each sample:

```{r}
samp.dist <- NULL
for(i in 1:1000){samp.dist <- c(samp.dist, mean(rexp(40, rate=0.2)))}
```   

Now, we plot this sampling distribution:  

```{r fig1chunk, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

#sampling distribution of sample means
g3 <- ggplot() 
g3+geom_histogram(aes(x=samp.dist, fill=..count..), col="black")+
        scale_fill_gradient("Count", 
                            low = "green", 
                            high = "red")+
        
        #mean of sampling dist.
        geom_vline(aes(xintercept=mean(samp.dist)),
                   linetype="longdash", 
                   col="blue")+
        geom_text(aes(x=mean(samp.dist)*1.15, 
                      y=100,
                      label=paste0("Mean of \n Sampling Distribution = ",
                                   round(mean(samp.dist), 4))
                      ), 
                  col="blue", size=3
                  )+
        #theoretical mean
        geom_vline(aes(xintercept=1/0.2),
                       linetype="longdash",
                       col="darkblue"
                   )+
        geom_text(aes(x=(1/0.2)*1.15, 
                      y=90,
                      label=paste0("Theoretical Mean of \n Exponential Distribution = ", 
                                   round(1/0.2, 4))
                      ), 
                  col="darkblue", size=3
                  )+
        
        #variance of sampling dist.
        geom_hline(aes(yintercept=var(samp.dist)*100),
                       linetype="longdash",
                       col="blue"
                   )+
        geom_text(aes(y=var(samp.dist)*110, 
                      x=3,
                      label=paste0("Variance of \n Sampling Distribution = ", 
                                   round(var(samp.dist), 4))
                      ), 
                  col="blue", size=3
                  )+
        #theoretical variance = population variance/n
        geom_hline(aes(yintercept=((1/(0.2^2))/40)*100),
                       linetype="longdash",
                       col="darkblue")+
        geom_text(aes(y=((1/(0.2^2))/40)*90, 
                      x=3,
                      label=paste0("Theoretical Variance \n of Distribution = ", 
                                   round(((1/(0.2^2))/40), 4))
                      ), 
                  col="darkblue", size=3
                  )+                 
        labs(title="Sampling Distribution of Sample Means \n of the Exponential Distribution (1000 Trials)",
                x="Means", y="Count")+
        theme_bw()+
        theme(legend.position="none")
```  

The distribution of a 1000 means of 40 random exponentials has begun to resemble a Gaussian distribution.   

### Comparing Theoretical Variance and Sample Mean

We continue to note from the earlier plot that the sampling distribution's mean and variance may be given by:
```{r echo=FALSE}
#distribution sample mean and variance
print(paste0("Mean of Sampling Distribution: ", round(mean(samp.dist), 4)))
print(paste0("Variance of Distribution: ", round(var(samp.dist), 4)))
```   

We noted earlier that the theoretical mean of the exponential distribution is given by $1/\lambda$ ($\lambda = 0.2$), which is equal to `r 1/0.2`. When we compare this to sampling distribution's mean = `r round(mean(samp.dist), 3)`, we see that it is a fairly good approximation.   

Furthermore, the theoretical variance of the sampling distribution is given by the variance of the population mean/sample size, or $(1/\lambda^{2})/n$, which = `r (1/0.2^2)/40`. Here, it seems that the variance obtained from the sampling distribution of `r round(var(samp.dist), 4)`, is a fair approximation of the theoretical variance of the sampling distribution obtained from the population mean.   

## Verifying Normality

In order to confirm that the sampling distribution obtained above is indeed normal, we will test a consequence of the central limit theorem, i.e. _as the sample size approaches infinity, the variation of the sampling distribution gets smaller and thus results in a closer approximation of the population mean_. We will conduct two simulations with larger sample sizes, n= 100 and n = 1000. We will conduct a 1000 trials and compute the variance and skew to give us an idea of the density curve.   

```{r}
#sim 1, sample size = 100
samp.dist1 <- NULL
for(i in 1:1000){samp.dist1 <- c(samp.dist1, mean(rexp(100, rate=0.2)))}

#sim 2, sample size=1000
samp.dist2 <- NULL
for(i in 1:1000){samp.dist2 <- c(samp.dist2, mean(rexp(1000, rate=0.2)))}
```  

```{r  fig2chunk, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
# PLOT SIM 1
g4 <- ggplot() + geom_histogram(aes(x=samp.dist1, y=..density.., fill=..count.., col="black"))+
        scale_fill_gradient("Count", low = "green", high = "red")+
        geom_density(aes(samp.dist1), col="black", size=1)+
        #mean of sampling dist.
        geom_vline(aes(xintercept=mean(samp.dist1)),linetype="longdash", col="blue")+
        geom_text(aes(x=mean(samp.dist1)*1.2, y=.9,label=paste0("Mean = ",
                                   round(mean(samp.dist1), 4))), col="blue", size=3)+
        #variance of sampling dist.
        geom_text(aes(y=.8, x=mean(samp.dist1)*1.2,
                      label=paste0("Variance = ", round(var(samp.dist1), 4))), col="blue", size=3)+ 
        #kurtosis
        geom_text(aes(y=.7,x=mean(samp.dist1)*1.2,
                      label=paste0("Kurtosis = ", round(kurtosis(samp.dist1), 4))),col="blue", size=3)+
        #skew
        geom_text(aes(y=.6,x=mean(samp.dist1)*1.2,
                      label=paste0("Skew = ", round(skewness(samp.dist1), 4))),col="blue", size=3)+
        labs(title="Sampling Distribution \n Sample Size = 100 \n 1000 Trials",x="Means", y="Density")+theme_bw()+theme(legend.position="none")

# PLOT SIM 2
g5 <- ggplot() + geom_histogram(aes(x=samp.dist2, y=..density.., fill=..count.., col="black"))+scale_fill_gradient("Count", low = "green", high = "red")+
        geom_density(aes(samp.dist2), col="black", size=1)+
        #mean of sampling dist.
        geom_vline(aes(xintercept=mean(samp.dist2)),linetype="longdash", col="blue")+
        geom_text(aes(x=mean(samp.dist2)*1.05, y=.9+1.5,
                      label=paste0("Mean  = ",round(mean(samp.dist2), 4))), col="blue", size=3)+
        #variance of sampling dist.
        geom_text(aes(y=.8+1.4, x=mean(samp.dist2)*1.05,
                      label=paste0("Variance = ", round(var(samp.dist2), 4))), col="blue", size=3)+ 
        #kurtosis
        geom_text(aes(y=.7+1.3,x=mean(samp.dist1)*1.05,
                      label=paste0("Kurtosis = ", round(kurtosis(samp.dist2), 4))),col="blue", size=3)+
        #skew
        geom_text(aes(y=.6+1.2,x=mean(samp.dist2)*1.05,
                      label=paste0("Skew = ", round(skewness(samp.dist2), 4))),col="blue", size=3)+
        labs(title="Sampling Distribution \n Sample Size = 1000 \n 1000 Trials",
                x="Means", y="Density")+theme_bw()+theme(legend.position="none")
grid.arrange(g4, g5, ncol=2)
```   

Consequently, we observe the following:  

* The mean is approximated more closely as the sample size increases  
* Variance of sampling distribution significantly decreases  
* The skew of the density curve gets smaller  

We can thus conclude that the sampling distribution of means of the exponential distribution approaches a normal distribution as the sample sizes increase.   

**END OF PART 1**  

# APPENDIX 1: GIT CODE

The entire markdown file can be found in this [github repo](https://github.com/ashirwad08/StatisticalInference.git).  

# APPENDIX 2: PLOT CODE 

## Comparing Variance and Mean to Theoretical Variance and Mean

```{r fig1chunk, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
```    

## Verifying Normality  

```{r fig2chunk, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
```


