---
title: "Exponential Distribution Simulation; Tooth Growth Analysis"
author: "Ash Chakraborty"
date: "Wednesday, June 17, 2015"
output:
  word_document:
    highlight: zenburn
  pdf_document: default
  html_document:
    theme: cerulean
---

# PART 1: SIMULATION - Sampling the Exponential Distribution

Part 1 of this report investigates the _Exponential Distribution_ and compares it to the _Central Limit Theorem_. We conduct simulations on this distribution with a sample size of 40 exponentials by generating a sampling distribution of sample means. The consequences of this sampling distribution will be evaluated for adherence to the central limit theorem.  

```{r global_options, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE, highlight=TRUE}
library(PerformanceAnalytics)
library(ggplot2)
library(gridExtra)
```
```{r echo=TRUE}
set.seed(123) 
sample <- rexp(40, rate=0.2)
```  
We conduct a 1000 simulations, and extract the mean of each sample:

```{r echo=TRUE}
samp.dist <- NULL
for(i in 1:1000) {samp.dist <- c(samp.dist, mean(rexp(40, rate=0.2)))}
```   
## Comparing Variance and Mean to Theoretical Variance and Mean  

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}

#sampling distribution of sample means
g3 <- ggplot() 
g3+geom_histogram(aes(x=samp.dist, fill=..count..), col="black")+
        scale_fill_gradient("Count", 
                            low = "green", 
                            high = "red")+
        
        #mean of sampling dist.
        geom_vline(aes(xintercept=mean(samp.dist)),
                   linetype="longdash", 
                   col="blue")+
        geom_text(aes(x=mean(samp.dist)*1.15, 
                      y=100,
                      label=paste0("Mean of \n Sampling Distribution = ",
                                   round(mean(samp.dist), 4))
                      ), 
                  col="blue", size=3
                  )+
        #theoretical mean
        geom_vline(aes(xintercept=1/0.2),
                       linetype="longdash",
                       col="darkblue"
                   )+
        geom_text(aes(x=(1/0.2)*1.15, 
                      y=90,
                      label=paste0("Theoretical Mean of \n Exponential Distribution = ", 
                                   round(1/0.2, 4))
                      ), 
                  col="darkblue", size=3
                  )+
        
        #variance of sampling dist.
        geom_hline(aes(yintercept=var(samp.dist)*100),
                       linetype="longdash",
                       col="blue"
                   )+
        geom_text(aes(y=var(samp.dist)*110, 
                      x=3,
                      label=paste0("Variance of \n Sampling Distribution = ", 
                                   round(var(samp.dist), 4))
                      ), 
                  col="blue", size=3
                  )+
        #theoretical variance = population variance/n
        geom_hline(aes(yintercept=((1/(0.2^2))/40)*100),
                       linetype="longdash",
                       col="darkblue")+
        geom_text(aes(y=((1/(0.2^2))/40)*90, 
                      x=3,
                      label=paste0("Theoretical Variance \n of Distribution = ", 
                                   round(((1/(0.2^2))/40), 4))
                      ), 
                  col="darkblue", size=3
                  )+                 
        labs(title="Sampling Distribution of Sample Means \n of the Exponential Distribution (1000 Trials)",
                x="Means", y="Count")+
        theme_bw()+
        theme(legend.position="none")
```  

The distribution has begun to resemble a Gaussian distribution.
```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
#distribution sample mean and variance
print(paste0("Mean of Sampling Distribution: ", round(mean(samp.dist), 4)))
print(paste0("Variance of Distribution: ", round(var(samp.dist), 4)))
```   
The theoretical mean of the exponential distribution is given by $1/\lambda$ ($\lambda = 0.2$) = `r 1/0.2`. **The sampling distribution's mean is a fairly good approximation**. The theoretical variance of the sampling distribution: $(1/\lambda^{2})/n$ = `r (1/0.2^2)/40`. **The variance obtained from the sampling distribution is a good approximation**.  

## Verifying Distribution Normality

We will conduct two simulations with larger sample sizes (n= 100 & 1000), conduct a 1000 trials and compute the _variance_ and _skew_ :

```{r echo=TRUE}
#sim 1, sample size = 100
samp.dist1 <- NULL
for(i in 1:1000){samp.dist1 <- c(samp.dist1, mean(rexp(100, rate=0.2)))}
#sim 2, sample size=1000
samp.dist2 <- NULL
for(i in 1:1000){samp.dist2 <- c(samp.dist2, mean(rexp(1000, rate=0.2)))}
```  
Plotting the resulting sampling distributions side by side:  

```{r eval= TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
# PLOT SIM 1
g4 <- ggplot() + geom_histogram(aes(x=samp.dist1, 
                                    y=..density.., 
                                    fill=..count.., 
                                    col="black"))+
        scale_fill_gradient("Count", 
                            low = "green", 
                            high = "red")+
        geom_density(aes(samp.dist1), col="black", size=1)+
        
        #mean of sampling dist.
        geom_vline(aes(xintercept=mean(samp.dist1)),
                   linetype="longdash", 
                   col="blue")+
        geom_text(aes(x=mean(samp.dist1)*1.2, 
                      y=.9,
                      label=paste0("Mean = ",
                                   round(mean(samp.dist1), 4))
                      ), 
                  col="blue", size=3
                  )+

        #variance of sampling dist.
        geom_text(aes(y=.8, 
                      x=mean(samp.dist1)*1.2,
                      label=paste0("Variance = ", 
                                   round(var(samp.dist1), 4))
                      ), 
                  col="blue", size=3
                  )+ 
        
        #kurtosis
        geom_text(aes(y=.7,
                      x=mean(samp.dist1)*1.2,
                      label=paste0("Kurtosis = ", round(kurtosis(samp.dist1), 4))
                      ),
                  col="blue", size=3
                  )+
        
        #skew
        geom_text(aes(y=.6,
                      x=mean(samp.dist1)*1.2,
                      label=paste0("Skew = ", round(skewness(samp.dist1), 4))
                      ),
                  col="blue", size=3
                  )+
        
        labs(title="Sampling Distribution \n Sample Size = 100 \n 1000 Trials",
                x="Means", y="Density")+
        theme_bw()+
        theme(legend.position="none")

# PLOT SIM 2
g5 <- ggplot() + geom_histogram(aes(x=samp.dist2, 
                                    y=..density.., 
                                    fill=..count.., 
                                    col="black"))+
        scale_fill_gradient("Count", 
                            low = "green", 
                            high = "red")+
        geom_density(aes(samp.dist2), col="black", size=1)+
        
        #mean of sampling dist.
        geom_vline(aes(xintercept=mean(samp.dist2)),
                   linetype="longdash", 
                   col="blue")+
        geom_text(aes(x=mean(samp.dist2)*1.05, 
                      y=.9+1.5,
                      label=paste0("Mean  = ",
                                   round(mean(samp.dist2), 4))
                      ), 
                  col="blue", size=3
                  )+

        #variance of sampling dist.
        geom_text(aes(y=.8+1.4, 
                      x=mean(samp.dist2)*1.05,
                      label=paste0("Variance = ", 
                                   round(var(samp.dist2), 4))
                      ), 
                  col="blue", size=3
                  )+ 
        
        #kurtosis
        geom_text(aes(y=.7+1.3,
                      x=mean(samp.dist1)*1.05,
                      label=paste0("Kurtosis = ", round(kurtosis(samp.dist2), 4))
                      ),
                  col="blue", size=3
                  )+
        
        #skew
        geom_text(aes(y=.6+1.2,
                      x=mean(samp.dist2)*1.05,
                      label=paste0("Skew = ", round(skewness(samp.dist2), 4))
                      ),
                  col="blue", size=3
                  )+
        
        labs(title="Sampling Distribution \n Sample Size = 1000 \n 1000 Trials",
                x="Means", y="Density")+
        theme_bw()+
        theme(legend.position="none")

grid.arrange(g4, g5, ncol=2)
```   

* The mean is approximated more closely as the sample size increases (in line with CLT)
* Variance of sampling distribution significantly decreases (in line with CLT)
* The skew of the density curve get smaller, suggesting a normal curve

**END OF PART 1**  

# PART 2: Analyzing Tooth Growth Data

We run some summary analysis on this dataset, in addition to testing hypotheses to extract any relationships between supplement, dosage, and tooth length.

## Data Summary

```{r echo=FALSE}
library(datasets)
#storing means separately to plot lines later
means <- with(ToothGrowth, tapply(len, supp, mean))
means.df <- data.frame(supp=levels(ToothGrowth$supp), means=c(means[[1]], means[[2]]))        
```   
The scatter below shows the dataset spread out into groups by supplement and dosage:   

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}

g1 <- ggplot(data=ToothGrowth, aes(x=dose, y=len, group=supp, col=supp, size=len) )
g1 + geom_point()+
        geom_jitter(position=position_jitter(width=0.1))+
        geom_smooth(alpha=0.3, method="loess")+ 
        geom_hline(data=means.df, aes(yintercept=c(means[[1]], means[[2]]), col=supp), linetype="longdash")+
        geom_text(data=means.df, aes(x=1.5, y=c(means[[1]], means[[2]]), col=supp, 
                                     label=c(paste0("Mean = ", round(means[[1]], 2)), 
                                             paste0("Mean = ", round(means[[2]], 2)))), 
                                             size=4, vjust=1)+
        labs(title="Scatter: Tooth Growth Length Vs. Dose\nof Supplements OJ and VC") +
        theme_bw() 
```   

* There is a positive correlation between length and dosage amount 0.5 and 1.0 for both supplements.
* There is _no data_ for either supplement at dosage amount 1.5.
* Any effect on length at dosage amount of 2.0 for either supplement is unclear.
* Overall, the average sample tooth growth length is greater for "OJ" than it is for "VC".

## Hypothesis Testing

In the first test (which serves as the template for the subsequent tests), we compare the effect on tooth growth between supplements OJ and VC at dose = 0.5. We **assume** that the NULL hypothesis, $H_0:\mu_{xOJ}=\mu_{xVC}$, is true; independent groups, and unequal variance.

```{r echo=TRUE}
# Function to take in the dosage group and perform a t-test on supplement effect on length
doseTest <- function(d) {
        testdf <- NULL
        testdf <- subset(ToothGrowth, dose==d, select=c('len', 'supp'))
        # apply t.test to compare the mean lengths by supplement at the dosage group
        t.test(len ~ supp, paired=FALSE, data=testdf)}
doseTest(0.5)
```
Here, we see that the 95% confidence interval is above zero. The p value of `r round(doseTest(0.5)$p.value*100, 2)`% is very unlikely, thus causing us to _reject_ the NULL hypothesis. **Under a similar assumption, we repeat the _t-test_ to get the probability of the alternate hypothesis for the other dosage groups**:
```{r echo=TRUE}
doseTest(1)$p.value
doseTest(2)$p.value
```   
We note that for dose = 1.0, the p value of `r round(doseTest(1)$p.value*100, 2)`% is unlikely, thus causing us to **reject** the NULL hypothesis. However, for dose = 2.0, the p value of `r round(doseTest(2)$p.value*100, 2)`% is well within the 95th percentile of values that comply with the NULL hypothesis. In this case, then, the NULL hypothesis **stands**.   

# Assumptions and Conclusions  

The following assumptions must be restated for the tests conducted above:  

* The subjects in each dosage groups are independent (not paired)
* Unequal variance between groups has been assumed  

The scatter plot and the tests performed on the dataset help us conclude the following:  

* At dose=0.5, OJ results in higher tooth growth than VC
* At dose=1.0, OJ results in higher tooth growth than VC
* At dose=2.0, there is no discernible difference in tooth growth between the supplements
* Overall, it seems that OJ is more effective than VC at increasing tooth growth when administered at doses 0.5 or 1.0.  

**END OF REPORT**  

# APPENDIX 1: GIT CODE

The entire markdown file can be found in this [github repo](https://github.com/ashirwad08/StatisticalInference.git).  

# APPENDIX 2: PLOT CODE BLOCKS 

## (PART 1) Comparing Variance and Mean to Theoretical Variance and Mean

```{r eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

#sampling distribution of sample means
g3 <- ggplot() 
g3+geom_histogram(aes(x=samp.dist, fill=..count..), col="black")+
        scale_fill_gradient("Count", 
                            low = "green", 
                            high = "red")+
        
        #mean of sampling dist.
        geom_vline(aes(xintercept=mean(samp.dist)),
                   linetype="longdash", 
                   col="blue")+
        geom_text(aes(x=mean(samp.dist)*1.15, 
                      y=100,
                      label=paste0("Mean of \n Sampling Distribution = ",
                                   round(mean(samp.dist), 4))
                      ), 
                  col="blue", size=3
                  )+
        #theoretical mean
        geom_vline(aes(xintercept=1/0.2),
                       linetype="longdash",
                       col="darkblue"
                   )+
        geom_text(aes(x=(1/0.2)*1.15, 
                      y=90,
                      label=paste0("Theoretical Mean of \n Exponential Distribution = ", 
                                   round(1/0.2, 4))
                      ), 
                  col="darkblue", size=3
                  )+
        
        #variance of sampling dist.
        geom_hline(aes(yintercept=var(samp.dist)*100),
                       linetype="longdash",
                       col="blue"
                   )+
        geom_text(aes(y=var(samp.dist)*110, 
                      x=3,
                      label=paste0("Variance of \n Sampling Distribution = ", 
                                   round(var(samp.dist), 4))
                      ), 
                  col="blue", size=3
                  )+
        #theoretical variance = population variance/n
        geom_hline(aes(yintercept=((1/(0.2^2))/40)*100),
                       linetype="longdash",
                       col="darkblue")+
        geom_text(aes(y=((1/(0.2^2))/40)*90, 
                      x=3,
                      label=paste0("Theoretical Variance \n of Distribution = ", 
                                   round(((1/(0.2^2))/40), 4))
                      ), 
                  col="darkblue", size=3
                  )+                 
        labs(title="Sampling Distribution of Sample Means \n of the Exponential Distribution (1000 Trials)",
                x="Means", y="Count")+
        theme_bw()+
        theme(legend.position="none")
```    

## (PART 1) Verifying Normality  

This code chunk is quite large and may be found in its entirety [in github: markdown](https://github.com/ashirwad08/StatisticalInference.git).  

## (PART 2) ToothGrowth Summary

```{r eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

g1 <- ggplot(data=ToothGrowth, aes(x=dose, y=len, group=supp, col=supp, size=len) )
g1 + geom_point()+
        geom_jitter(position=position_jitter(width=0.1))+
        geom_smooth(alpha=0.3, method="loess")+ 
        geom_hline(data=means.df, aes(yintercept=c(means[[1]], means[[2]]), col=supp), linetype="longdash")+
        geom_text(data=means.df, aes(x=1.5, y=c(means[[1]], means[[2]]), col=supp, 
                                     label=c(paste0("Mean = ", round(means[[1]], 2)), 
                                             paste0("Mean = ", round(means[[2]], 2)))), 
                                             size=4, vjust=1)+
        labs(title="Scatter: Tooth Growth Length Vs. Dose\nof Supplements OJ and VC") +
        theme_bw() 
```   

